Iterative Monomorphisation

### 2024-02-21 iterative_monomorphisation.tex
couple of small changes

@@

% TODO, at the end check the following
%   - formulas -> formulae
%   - clause -> formula
%   - monomorphization -> monomorphisation
%   - higher-order -> higher order
%   - TH1 with ugly 1
%   - correctly spelled TPTP, Leo-III, Zipperposition, E, Vampire
%   - use of formula vs clause
%   - polymorphic and monomorphic
%   - title of bits of the algorithms
%   - eg or ex -> e.g.
%   - function symbol -> symbol instance (depends on context)
%   - ... -> \dots
%   - comma splices!!!!
%   - ``'' -> `'
%   - use \Phi instead of F for formulae
%   - concrete types -> monomorphic types
%   - use in not \in when we are not precisely dealing with set membership
%   - $\Phi$ for problem and $\varphi_1$ etc.
%   - n-uples -> n-tuples
%   - ize -> ise (in cas i forgot any)
%   - remove vertical space before lists (if allowed)
%   - add full stops to end of sentences in itemize
%   - B\"ome -> B\"ohme

@@

  % apps are rank-1 polymorphic, but provers are mono
  % there's a gap
  % one solution is to extend the provers to support polymorphism;
  % this was done e.g. for Vampire \cite{xxx}
  % but this is a lot of work, and needs to be done for every prover

@@

 % * the alternative is to translate polymorphic problems to monomorphic problems
 %   * two main approaches:
 %     * complete encoding of polymorphism, using guards or tags,
 %       as in Blanchette et al.~\cite{mono-trans} and Bobot et al.~\cite{expr-poly-types}
 %     * incomplete encoding of polymorphism based on iterative monomorphisation
 %       such as introduced by Böhme \cite[Section 2.2.1]{sb-phd}, whereby type
 %       variables are heuristically instantiated by concrete types

@@

  %* By a type version of compactness, if a set of polymorphic
  %  formulae is unsatisfiable, then there exists a finite set of monomorphic
  %  instances of the formulae that is also unsatisfiable. However, this
  %  finite set cannot be computed \cite[Section 2, Theorem 1]{expr-poly-types}; hence any approach based on
  %  instantiation of type variables is incomplete.

@@

% * Böhme's iterative approach is implemented as part of the SMT integration \cite[Chapter 2]{sb-phd}
%     in Isabelle/HOL. This implementation is also used by
%    Sledgehammer \cite{judgement, hammer} to interface automatic theorem provers based on
%    superposition.
%
%  * The main drawback of Böhme's approach is that it is only briefly documented,
%    as a single paragraph in his PhD thesis \cite[Section 2.2.1]{sb-phd}
%  * In this paper, we present an algorithm inspired by his description and
%    implementation (Section~\ref{sec:high level-algorithm}).
%    * we aimed at mathematical precision, so that people who want to reimplement
%      the algorithm can follow this blueprint
%    * we also show how to bound the algorithm to avoid explosions in practice
%      (Section~\ref{sec:low-level-algorithm}).
%      ``the explosive nature of the procedures requires various bounds which
%      limit the number of type arguments and clauses that are generated.''

@@

%These iterations are required to account for $n$-ary type constructors.
Consider the unary type constructor \ty{list}. If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, it may be possible to generate the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc. However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has already been generated. This example also makes it clear that the set of instances that can be generated is infinite.

@@

Our empirical evaluation on TPTP \cite{tptp} problems attempts to answer three questions (Section~\ref{sec:evaluation}):
%
\begin{enumerate}
\item Is the new Zipperposition with the E backend more successful on polymorphic problems than Zipperposition without backend?

@@

Our findings are as follows:
%
\begin{enumerate}
\item Zipperposition benefits substantially from the E backend.

@@

This algorithm works independently of the structure of the problem's formulae. It relies exclusively on the formulae's monomorphic and polymorphic symbol instances. Type variables are assumed to be implicitly universally quantified at a formula's top level. The precise form of formulae is left unspecified.
%
Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic. In particular, it can operate in the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.

%The following terms are used in the algorithm descriptions.

%{\obeylines
%
%\noindent\kern\parindent
%  * algorithm works a priori with any rank-1 polymorphic logic
%
%  * in particular, it works with polymorphic first and higher order logic as 
%    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}
%
%  * the algorithm does not need to inspect the precise structure of formulae
%  * instead, it works on monomorphic or polymorphic instances of symbols that 
%    occur in the formulae
%  * type quantification is implicit, meaning that all type variables occurring
%    in a formula are assumed to be universally quantified at the formula's top
%    level
%}

% TH1 ugly for height of 1 reasons, can be fixed with case feature

@@

\begin{definition}%[Type]
A (\emph{polymorphic}) \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\), \(\ty{map}(\ty{int},\ty{string})\)).
If $n = 0$, we omit the parentheses (e.g.\ \(\ty{int}\)).
\end{definition}

\begin{definition}%[Monomorphism and polymorphism]
A type is \emph{monomorphic} if it contains no type variables.
\end{definition}

\begin{definition}%[Function symbol]
A (function or predicate) {symbol} \(f\) has a \emph{type arity} that defines the number of type arguments it takes. A \emph{symbol instance} is a symbol applied to type arguments listed between angle brackets: \(f\langle \tau_1, \dots, \tau_n\rangle\), where each $\tau_i$ is a type. If $n = 0$, we omit the brackets (e.g.\ $f$).
\end{definition}

@@

%We start with a high level description of the iterative monomorphisation algorithm.
The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem. It operates by applying an arbitrary number of iterations. Each iteration takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae. A single iteration consists of a collection phase
% (Section~\ref{ssec:collection})
and an instantiation phase.
% (Section~\ref{ssec:instantiation})
Once the iterations are completed, a final step filters out all non-monomorphic formulae returned by the last iteration.

%\subsection{Collection}
%\label{ssec:collection}

@@

%\subsection{Instantiation}
%\label{ssec:instantiation}

@@

%\subsection{Example}

@@

%
%
%On the first iteration, two matchings are attempted, that of \(\alpha\) against \(\ty{int}\) which yields \(\sigma = \{\alpha \mapsto \ty{int}\}\) and that of \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails. Then, \(\sigma\) is applied to [1], resulting in a new formula: \(\forall a: \ty{int}, \mathit{as}:\ty{list}(\ty{int}), \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
%
%At the beginning of the second iteration, the clause generated in the previous iteration is used to update \(M(f)\), adding \((\ty{list}(\ty{int}))\). \(P\) has no new elements. Similarly to the first iteration we attempt the following matchings:
%\begin{itemize}
%   \item \(\alpha\) against \(\ty{int}\) which succeeds resulting in \(\sigma\)
%   \item \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails
%   \item \(\alpha\) against \(\ty{list}(\ty{int})\) resulting in \(\sigma'=\{\alpha\mapsto\ty{list}(\ty{int})\}\)
%   \item \(\ty{list}(\alpha)\) against \(\ty{list}(\ty{int})\) resulting in \(\sigma\)
%\end{itemize}
%
%From these substitutions, a single new formula can be generated from the application of \(\sigma'\) to [1], giving us: \(\sigma'(\text{[1]}) = \forall a: \ty{list}(\ty{int}), \mathit{as}:\ty{list}(\ty{list}(\ty{int})), \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
%
%This new formula will be used to add \((\ty{list}(\ty{list}(\ty{int})))\) to \(M(f)\) at the beginning of the third iteration. Similarly to the previous iteration, a single new formula will be generated. It is obtained by applying the new substitution \(\sigma'' = \{\alpha\mapsto\ty{list}(\ty{list}(\ty{int}))\}\) to [1].
\end{example}

@@

The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a more low level algorithm with the following features.
%
\begin{enumerate}
    \item[\labelitemi] Bounds are used at several stages of the algorithm to curb the highly explosive enumerations.

@@

\begin{quote}
\begin{algorithm}[H]
   \SetAlgoNoLine
%   \LinesNumbered
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
   \SetKwData{Continue}{continue}\SetKwData{SubstRes}{substitution\_result}
   \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
   \SetKwFunction{Match}{match}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

   \Input{\,\(\varphi\), \(M\), \(P\)}
   \Output{\,a set of substitutions}
   \BlankLine

   \(S\leftarrow \emptyset\)\;
   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
         \(\sigma\leftarrow \{\}\)\;
         \Continue \(\leftarrow\) \True\;
         \(i \leftarrow 0\)\;
         \While{\(i \leq n\) \And \Continue}{
            \SubstRes \(\leftarrow\) \Match{\(\upsilon_i\), \(\tau_i\)}\;
            \lIf{\SubstRes = \Success{\(\sigma_i\)} \And \Compatible{\(\sigma\), \(\sigma_i\)}}{
               \(\sigma\leftarrow\sigma_i\circ\sigma\) 
            }
            \lElse{\Continue\(\leftarrow\)\False}
            \(i\leftarrow i+1\)\;
         }
         \lIf{\Continue}{\(S\leftarrow S\cup \{\sigma\}\)}
      }

>>>>

\definecolor{grey}{rgb}{0.8,0.8,0.8}
\definecolor{algoColorKeyword}{named}{grey}
\makeatletter
% Block with a vertical line
\renewcommand{\algocf@Vsline}[1]{%
   \strut\par\nointerlineskip%
   \algocf@push{\skiprule}%
   \hbox{%
      {\color{algoColorKeyword}\vrule}
      \vtop{\algocf@push{\skiptext}%
      \vtop{\algocf@addskiptotal\advance\hsize by -\skiplength #1}}%
   }%
   \algocf@pop{\skiprule}%
}

@@

% TODO, at the end check the following
%   - formulas -> formulae
%   - clause -> formula
%   - monomorphization -> monomorphisation
%   - higher-order -> higher order
%   - TH1 with ugly 1
%   - correctly spelled TPTP, Leo-III, Zipperposition, E, Vampire
%   - use of formula vs clause
%   - polymorphic and monomorphic
%   - title of bits of the algorithms
%   - TH1 ugly for height of 1 reasons, can be fixed with case feature
%   - eg or ex -> e.g.
%   - function symbol -> symbol instance (depends on context)
%   - ... -> \dots
%   - comma splices!!!!
%   - ``'' -> `'
%   - use \Phi instead of F for formulae
%   - concrete types -> monomorphic types
%   - use in not \in when we are not precisely dealing with set membership
%   - $\Phi$ for problem and $\varphi_1$ etc.
%   - n-uples -> n-tuples
%   - ize -> ise (in cas i forgot any)
%   - remove vertical space before lists (if allowed)
%   - add full stops to end of sentences in itemize
%   - B\"ome -> B\"ohme

@@

@@

@@

@@

@@

Consider the unary type constructor \ty{list}. If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, it may be possible to generate the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc. However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has already been generated. This example also makes it clear that the set of instances that can be generated is infinite.

@@

Our empirical evaluation on TPTP \cite{tptp} problems attempts to answer three questions (Section~\ref{sec:evaluation}):
\begin{enumerate}
\item Is the new Zipperposition with the E backend more successful on polymorphic problems than Zipperposition without backend?

@@

Our findings are as follows:
\begin{enumerate}
\item Zipperposition benefits substantially from the E backend.

@@

This algorithm works independently of the structure of the problem's formulae. It relies exclusively on the formulae's monomorphic and polymorphic symbol instances. Type variables are assumed to be implicitly universally quantified at a formula's top level. The precise form of formulae is left unspecified.
Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic. In particular, it can operate in the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.

@@

\begin{definition}
A (\emph{polymorphic}) \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\), \(\ty{map}(\ty{int},\ty{string})\)).
If $n = 0$, we omit the parentheses (e.g.\ \(\ty{int}\)).
\end{definition}

\begin{definition}
A type is \emph{monomorphic} if it contains no type variables.
\end{definition}

\begin{definition}
A (function or predicate) {symbol} \(f\) has a \emph{type arity} that defines the number of type arguments it takes. A \emph{symbol instance} is a symbol applied to type arguments listed between angle brackets: \(f\langle \tau_1, \dots, \tau_n\rangle\), where each $\tau_i$ is a type. If $n = 0$, we omit the brackets (e.g.\ $f$).
\end{definition}

@@

The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem. It operates by applying an arbitrary number of iterations. Each iteration takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae. A single iteration consists of a collection phase and an instantiation phase. Once the iterations are completed, a final step filters out all non-monomorphic formulae returned by the last iteration.

@@

@@

@@

\end{example}

@@

The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a more low level algorithm with the following features.
\begin{enumerate}
    \item[\labelitemi] Bounds are used at several stages of the algorithm to curb the highly explosive enumerations.

@@

\begin{quote}
\begin{algorithm}[H]
   \caption{Substitution Generation}
%   \SetAlgoNoLine
%   \LinesNumbered
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
   \SetKwData{Continue}{continue}\SetKwData{SubstRes}{substitution\_result}
   \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
   \SetKwFunction{Match}{match}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

   \Input{\,formula \(\varphi\), monomorphic type argument map \(M\), non-monomorphic type argument map \(P\)}
   \Output{\,set of substitutions}
   \BlankLine

   \(S\leftarrow \emptyset\)\;
   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
         \(\sigma\leftarrow \{\}\)\;
         \Continue \(\leftarrow\) \True\;
         \(i \leftarrow 0\)\;
         \While{\(i \leq n\) \And \Continue}{
            \SubstRes \(\leftarrow\) \Match{\(\upsilon_i\), \(\tau_i\)}\;
            \Case{\SubstRes = \Success{\(\sigma_i\)}}{
               \lIf{\Compatible{\(\sigma\), \(\sigma_i\)}}{
                  \(\sigma\leftarrow\sigma_i\circ\sigma\)}
               \lElse{\Continue\(\leftarrow\)\False}
            }
            \lOther{\Continue\(\leftarrow\)\False}

         }
         \(i\leftarrow i+1\)\;

         \lIf{\Continue}{\(S\leftarrow S\cup \{\sigma\}\)}
      }
